Supplementary Information

### 2021-03-23 doc/manuscript/ForestGEO_treerings.Rmd
NP comments- intro

*(cut most of this paragraph? but JB really liked it):*
Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the long-term climate sensitivity of individual trees or $ANPP_{woody}$.
Traditional methods begin by fitting a function (commonly an exponential curve or a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous growing season [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean slope of the relationship among individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@pederson_framework_2020].
Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
In particular, analysis of species-level chronologies does not allow the direct characterization of known variation in climate sensitivity in relation to tree height, canopy position, or microhabitat characteristics such as topographic wetness index [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].

*(reword to ecology focus):*
Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
Over time frames of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad climatic gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided (e.g., by sampling at sites with characteristics that result in trees being more climate-limited), and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
*Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon. Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.*
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**

*(keep mostly the same):*
Tree diameter at breast height (DBH) scales predictably with numerous traits affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
Although climate correlations can be transformed to climate sensitivity [i.e., magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
For example, drought sensitivity tends to increase with tree size [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
To use tree rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].

>>>>

*(cut most of this paragraph? but JB really liked it):*
Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the long-term climate sensitivity of individual trees or $ANPP_{woody}$.
Traditional methods begin by fitting a function (commonly an exponential curve or a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous growing season [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean slope of the relationship among individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@carrer_individualistic_2011; @pederson_framework_2020].
Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
In particular, analysis of species-level chronologies does not allow the direct characterization of known variation in climate sensitivity in relation to tree height, canopy position, or microhabitat characteristics such as topographic wetness index [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].

*(reword to ecology focus):*
Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
Over time frames of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad climatic gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically not the first focus for sampling. 
Many worked around the non-linearity issues by sampling at sites with characteristics that result in trees being more climate-limited, though not all [@cook_treering_1982].
Importantly, traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
*Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon. Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.*
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**

*(keep mostly the same):*
Tree diameter at breast height (DBH) scales predictably with numerous traits affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence, *along with the influence of stand dynamics in closed-canopy forests *, is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only trees in canopy positions or those that appear to be old [@fritts_tree_1976; @huckaby_identification_2003; @pederson_external_2010].
While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
Although climate correlations can be transformed to climate sensitivity [i.e., magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
For example, drought sensitivity tends to increase with tree size [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size or canopy position [@rossi_agedependent_2007; @rollinson_climate_2021].
However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
To use tree rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
